Characterizing subarctic biomes for land surface modeling of pollution and climate risk.

### 2021-02-08 VegOzone.tex
Change figure
Change title and rewrite abstract.

%\title{The 2018 heatwave and its implications on ozone induced damage on vegetation in a subarctic climate}
\title{Ozone induced damage on vegetation in a subarctic climate: implications of Plant Functional Types on Phototoxic Ozone Dose modeling}

@@

\begin{abstract}
  In the future, persitent heatwaves may occur more often. This
  An increased occurrence of persistent heatwaves, as one possible future scenario, generates favorable conditions for the formation of ambient air ozone. Vegetation highly specialized to sub-arctic climate is vulnerable to rapid environmental changes inflicted by global warming and might become more susceptible to ozone in the future. Over large parts of Europe the summer 2018 had been extraordinarily hot and dry and caused large wildfires in northern Sweden in particular. This can be regarded as a test case for such a future scenario. In both 2018 and 2019, we have monitored ambient air ozone concentrations at the Norwegian Institute of Bioeconomy Research (NIBIO) Environment Centre Svanhovd in Northern Norway. Due to data acquisition problems, ozone concentrations for two weeks in July 2018 were missing from our record. We present a reconstruction based on probability density function with respect to the Swedish and Finnish atmospheric monitoring sites in the region. Over all, ozone concentrations did not differ significantly between the two years. While temperatures and global irradiance diverged significantly from multi annual mean, precipitation varied only to some extend. Coincidentally, we have observed ozone-induced visible injuries on clovers in the ozone garden at Svanhovd in 2018, but not in 2019. We investigate the difference in uptake of ozone using the DO3SE model, with respect to the typical vegetation (beech/birch and conifers) at the location. We assess whether critical levels for these species were breached during the summer seasons. We find that an unadjusted transfer of currently used standard parameters and methodes on ozone damage assessment (IPC Mapping Manual) to vegetation in the subarctics will result in an missinterpretation of POD1 values. {\bf TODO: Results. Difference in POD1 compared to mapping manual parameterization.}
\end{abstract}

@@

\begin{figure}[t]
  \includegraphics[width=8.3cm]{javis_funcs_temp_hist_grassland}\\
  \includegraphics[width=8.3cm]{javis_funcs_f_light_grassland}
\caption{TEXT}
\label{fig:f_temp_grassland}
\end{figure}

>>>>

%\title{The 2018 heatwave and its implications on ozone induced damage on vegetation in a subarctic climate}
\title{Characterizing subarctic biomes for land surface modeling of pollution and climate risk.}

@@

\begin{abstract}
Here we assess the importance of characterization of key meteorological variables and air pollution profiles on the modeled susceptibility of subarctic vegetation to ozone. Changes in climate (e.g. temperature, water availability) will manifest in both direct and indirect effects influencing the native species sensitivity to air pollution. A likely increase in frequency of persistent heatwaves in the future, will inflict heat related stress while generating favorable conditions for the formation of ambient air ozone, e.g. due to wildfires.
We have monitored ambient air ozone concentrations at the Norwegian Institute of Bioeconomy Research (NIBIO) Environment Centre Svanhovd in Northern Norway in the vegetation period 2018/19. Due to data acquisition problems, ozone concentrations for two weeks in July 2018 were missing from our record. We present a reconstruction based on probability density function with respect to the Swedish and Finnish atmospheric monitoring sites and compare with available surface ozone products for the region. Over all, ozone concentrations did not differ significantly between the two years. While temperatures and global irradiance diverged substantially from the multi annual mean precipitation varied only to some extend. 
We investigate the difference in uptake of ozone using the DO3SE model, for typical vegetation at the location. We find that bespoke parameterization for subarctic vegetation is essential for an accurate estimation of Phototoxic Ozone Dose (POD) and an assessment of damage risk. Accurate estimates of the start and length of growing season are also crucial.
We find that temperature and its derivative vapor pressure deficit are likely the most important climate variables while the variability of ozone concentration within climatologically reasonable limits plays a minor role.
The use of subarctic parameterization suggests an improvement in risk assessment as POD values are increased in 2018; corroborated by evidence of visible injury in the Svanhovd ozone garden.
\end{abstract}

@@

\begin{figure}[t]
  \includegraphics[width=8.3cm]{javis_funcs_temp_hist_grassland}\\
  \includegraphics[width=8.3cm]{javis_funcs_rad_hist_grassland}
\caption{TEXT}
\label{fig:f_temp_grassland}
\end{figure}
